---
title: "Stimulus Flowchart"
format:
  html:
    page-layout: full
    fig-responsive: false
    fig-align: center
toc: true
code-tools: true
editor: source
---

```{r}
#| message: false
#| warning: false
library(rPraat)
library(sosprosody)
library(tidyverse)
library(targets)
library(here)
tar_dependency <- tar_read(render_EDA)
```

## Recording and Setup


```{mermaid}
flowchart TD
classDef praat fill:#E46E85,stroke:#85144b,color:#ffffff;
classDef python fill:#366994,stroke:#FFC331,color:#FFC331;
classDef rstats fill:#6FA1D0,stroke:#245AAB,color:#ffffff;
classDef shell fill:#111111,stroke:#2ECC40,color:#ffffff;
%% Files to make
TriggerScript[[trigger2tg.praat]]
ExtractTakes[[extract_takes.praat]]
MFA[[force_align.sh]]
PitchTierScript[[gen_manips_pitchtier.praat]]
PraatSauce[[praatsauce.praat]]
SpectralAnalysis[[spectral_measures_analysis.qmd]]
MidPhonFigs[[midphon_figures.qmd]]

class TriggerScript praat
class ExtractTakes praat
class MFA shell
class PitchTierScript praat
class PraatSauce praat
class SpectralAnalysis rstats
class MidPhonFigs rstats


%% Define Directories
RawRec["📂00: Recordings"]
MonoRec["📂01: Mono Recordings"]
PossibleRecs["📂02: Possible Recordings"]
Textgrids["📂02a: Textgrids for Recordings"]
PitchTiers["📂02b: PitchTiers for Recordings"]



RawRec -- Extract mono channel --> TriggerScript --> MonoRec
    MonoRec -- Annotate textgrid to denote OK takes --> ExtractTakes
    ExtractTakes --> PossibleRecs
    PossibleRecs --> MFA -- Standardized textgrid annotations --> Textgrids
    PossibleRecs --> PitchTierScript --> PitchTiers
    
PossibleRecs -- Extract Spectral Measures --> PraatSauce
  PraatSauce --> SpectralAnalysis --> MidPhonFigs
    
```


This portion of the process has most of the manual labor.
After the recordings are done, I use the `trigger2tg.praat` script to segment
the individual target utterances, which contain ***many takes*** of the same sentence.
Then, I listen to all the takes and denote which ones are OK enough to 
possibly use later, which are marked on a TextGrid with the following format:[^tgformat]

[^tgformat]: I wrote these custom print method for TextGrids and PitchTiers
as part of the `sosprosody` package, available on my github at [tsostarics/sosprosody](https://github.com/tsostarics/sosprosody)

```{r}
tg.read("01_Mono/exp1_raw_3syl_01_HLL.TextGrid")
```

The `extract_takes.praat` script automatically extracts all of the takes
annotated on the second tier above, saving separate `.wav` files and TextGrids.
These textgrids have a single tier/interval containing the utterance, which
is fed into the force alignment step.

```{r}
tg.read("02_PossibleRecordings/grandmother_01_LHH_001.TextGrid")
```

### Force-aligned TextGrids and PitchTiers

The force alignment step uses the Montreal Force Aligner with the dictionary
and acoustic model as specified in the commands below.
Log files are saved for the validation and align procedures.
Note that there's some new out of vocabular words that were added to the
dictionary, see the `README` for details.

```{r}
read_lines(tar_read(force_align_script))
```

The resulting text grids have the following standardized format:

```{r}
tg.read("02_PossibleRecordings/MFA_textgrids/grandmother_01_LHH_001.TextGrid")
```

To make corresponding PitchTier files, the `gen_manips_pitchtier.praat` script
is used. The pitchtiers look like this:

```{r}
pt.read("02_PossibleRecordings/PitchTiers/grandmother_01_LHH_001.PitchTier")
```

## Selecting recordings

```{mermaid}
%%| fig-align: center
flowchart TD
classDef praat fill:#E46E85,stroke:#85144b,color:#ffffff;
classDef python fill:#366994,stroke:#FFC331,color:#FFC331;
classDef rstats fill:#6FA1D0,stroke:#245AAB,color:#ffffff;
classDef shell fill:#111111,stroke:#2ECC40,color:#ffffff;
classDef excel fill:#17874E,stroke:#134130,color:#ffffff;

RawEDA[[raw_recordings_eda.qmd]]
SylHelper[[syllable_helpers.R]]
SylSpecs[[syllabification.csv]]

class RawEDA rstats
class PromoScript python
class GetDurTiers rstats
class DurationScript praat
class DurationScript2 praat
class SylHelper rstats
class TierHelper rstats
class SylSpecs excel

Textgrids["📂02a: Textgrids for Recordings"]
PitchTiers["📂02b: PitchTiers for Recordings"]
Picked["📂03: Chosen Recordings"]
PickedTG["📂03a: Chosen TextGrids"]
FixedTG["📂03b: Fixed TextGrids"]
EDA(Summary statistics & EDA with raw files)
%%    PlotHelpers --> RawEDA
    Textgrids & PitchTiers  --> RawEDA --> EDA
    EDA -- Match recordings, pick best exemplars --> Picked
    Picked -.- PickedTG
    SylSpecs & PickedTG --> SylHelper
    SylHelper -- Add tiers, adjust boundaries --> FixedTG
```


The next part of the process has three contributions.
First, we can get an exploratory look at the recordings to get an exploratory
look at the distribution of recordings.
Second, we can decide which recordings are the best exemplars to use for later
steps of this process.
Third, we use the results of the exploratory analysis to decide on the durations to use for standardizing the nuclear tune length.

### Averaging process

This flowchart expands on how the durations are calculated.
This is done by first calculating the average duration for each utterance-tune pairing
from all takes of that pairing.
Second, we calculate the average duration of an utterance based on its average
duration in each tune condition; this is done to avoid skewing averages towards
a tune that had more recordings than the other tunes.
Once we have the average durations for each utterance, we calculate how
much each recording of that utterance would need to be scaled to reach the
average duration.
A file for each utterance-tune pairing is selected for further manipulation
based on which take is closest (i.e., requires the least amount of scaling)
to reach the target duration.

The textgrids for these recordings (~64) are manually checked to verify the
output of the MFA with particular attention for the boundaries of phones
on a syllable or word boundary.
These TextGrids have the following format, note the addition of nuclear
and syllable interval tiers:


```{r}
tg.read(here::here("03_ChosenRecordings/FixedTextGrids/branning_01_HLL_002.TextGrid"), 
        'auto')
```

::: {.column-screen-inset-right}

```{mermaid}
flowchart LR
classDef rstats fill:#6FA1D0,stroke:#245AAB,color:#ffffff;
GetDurTiers[[make_duration_tiers.R]]
class GetDurTiers rstats


class PromoScript python
    subgraph HLL[HLL Takes]
        direction LR
        HLL:1 -.- HLL:... -.- HLL:n
    end
    muHLL[HLL Average]
    HLL --> muHLL
    subgraph LHH[LHH Takes]
        direction LR
        LHH:1 -.- LHH:... -.- LHH:n
    end
    muLHH[LHH Average]
    LHH --> muLHH
    subgraph LHSLL[L+H*LL Takes]
        direction LR
        LHSLL:1 -.- LHSLL:... -.- LHSLL:n
    end
    muLHSLL[LHSLL Average]
    LHSLL --> muLHSLL
    subgraph LSHLL[L*+HLL Takes]
        direction LR
        LSHLL:1 -.- LSHLL:... -.- LSHLL:n
    end
    muLSHLL[LSHLL Average]
    LSHLL --> muLSHLL
    muHLL & muLHH & muLHSLL & muLSHLL --> average[Averaged Duration]
    average --> GetDurTiers
```

:::

## Duration manipulations

::: {.column-screen-inset-right}

```{mermaid}
flowchart TD
classDef praat fill:#E46E85,stroke:#85144b,color:#ffffff;
classDef python fill:#366994,stroke:#FFC331,color:#FFC331;
classDef rstats fill:#6FA1D0,stroke:#245AAB,color:#ffffff;
classDef shell fill:#111111,stroke:#2ECC40,color:#ffffff;
classDef excel fill:#17874E,stroke:#134130,color:#ffffff;

DurationScript[[apply_dt_manipulation.praat]]
DurationScript2[[apply_dt_manipulation.praat]]
TierHelper[[tiermanip_helpers.R]]

class RawEDA rstats
class PromoScript python
class GetDurTiers rstats
class DurationScript praat
class DurationScript2 praat
class SylHelper rstats
class TierHelper rstats
class SylSpecs excel

Picked["📂03: Chosen Recordings"]
FixedTG["📂03b: Fixed TextGrids"]
DurTiers["📂03c: Duration Tiers"]
AvgFiles["📂04: Scaled Recordings"]
DurTextGrids["📂04a: Duration-morphed TextGrids"]
NudgedFiles["📂05: Nudged Recordings"]
NudgedTextGrids["📂05b: Nudged TextGrids"]
NudgedDurationTiers["📂05a: Nudged DurationTiers"]

EDA(Summary statistics & EDA with raw files)
%%    PlotHelpers --> RawEDA
    FixedTG --> TierHelper
    EDA -- Determine durations to use --> TierHelper
    TierHelper --> DurTiers & DurTextGrids
    DurTextGrids -.- AvgFiles 
    DurTiers & Picked --> DurationScript --> AvgFiles
    TierHelper --> NudgedDurationTiers
    AvgFiles & NudgedDurationTiers -- Reduce final syllable duration --> DurationScript2
    DurationScript2 --> NudgedFiles & NudgedTextGrids
```

:::

Once the chosen recordings are selected and their textgrids checked, a second
round of averaging on the syllable level is done.
Here, the average duration of each syllable is calculated.
This is done separately for two- and three-syllable word sets.
These averages serve as target syllable durations for each word.
Duration Tiers are written for each file to scale the duration of each syllable
of the nuclear word to the target values.

Duration Tiers technically have the same exact specification as Pitch Tiers,
so I'll show an example of what these duration tiers look like by treating it
as a Pitch Tier.
The Y-axis isn't perfect because the range of the values is very small (both endpoints round to 1) but below you can see the two rows of points on the ends are at 1, but there's a region at 1.3 (appears as much higher) and a region at .94 (appears as slightly lower than 1).
Thus, we see that for this recording, the first syllable needed to be stretched while the second syllable needed to be shrunk slightly.

```{r}
durationtier <- 
  dt.read(here::here("03_ChosenRecordings/DurationTiers/branning_01_HLL_002.DurationTier"))

class(durationtier)['type'] <- "PitchTier"

durationtier
```

So far we have standardized the duration of the syllables using manually
calculated duration tiers given target values from the average durations across
utterances.
A second duration manipulation is also applied where the final syllable in each
word is shrunk by a constant value, such as .80 or .90. 
Because the syllables are already standardized, this value is applied in the
same way to every file.
This is done to correct for final lengthening that makes straight-line 
interpolation between the pitch accent and boundary tone targets sound a bit off.
The value to shrink by is largely arbitrary, and we try reducing the duration 
of the final syllable in steps of 5%, from 100% (original duration) to 65% of
the original duration.
When listening to the final resynthesis, it seems like 85% or 80% works best.

## Resynthesis

::: {.column-screen-inset-right}

```{mermaid}

flowchart TB
classDef praat fill:#E46E85,stroke:#85144b,color:#ffffff;
classDef python fill:#366994,stroke:#FFC331,color:#FFC331;
classDef rstats fill:#6FA1D0,stroke:#245AAB,color:#ffffff;
classDef shell fill:#111111,stroke:#2ECC40,color:#ffffff;

Exp1ResynthScript[[resynthesize_exp1.praat]]
Exp2ResynthScript[[resynthesize_exp2.praat]]
Exp3ResynthScript[[resynthesize_straight_alignment_continua.praat]]
Exp3cResynthScript[[resynthesize_exp3_curved.praat]]
Exp1dResynthScript[[resynthesize_monotonal_bottomout.praat]]
Exp2dResynthScript[[resynthesize_LHS_bottomout.praat]]
Exp3dResynthScript[[resynthesize_bitonal_bottomout.praat]]
ResynthAnalysis[[resynthesis_analysis.qmd]]
BezierHelpers[[bezier_helpers.R]]
TcogFigures[[tcog_figures.qmd]]


class Exp1ResynthScript praat
class Exp2ResynthScript praat
class Exp3ResynthScript praat
class Exp3cResynthScript praat
class Exp1dResynthScript praat
class Exp2dResynthScript praat
class Exp3dResynthScript praat
class ResynthAnalysis rstats
class BezierHelpers rstats
class TcogFigures rstats


AvgFiles["📂05: Nudged Recordings"]
Exp1Resynth["📂06: Exp1"]
Exp2Resynth["📂06b: Exp2"]
Exp3Resynth["📂06c: Exp3"]
Exp3cResynth["📂06cCurved: Exp3 Curved"]
Exp1dResynth["📂06d: Monotonal Bottomout"]
Exp2dResynth["📂06d: LHS Curved Bottomout"]
Exp3dResynth["📂06d: Bitonal Curved Bottomout"]
FixedFiles["📂07: Reduced Recordings"]
 
subgraph Resynthesis[Resynthesis]
  direction LR
  subgraph ResynthScripts[Resynthesis Scripts]
    direction TB
    Exp1ResynthScript ---
      Exp2ResynthScript ---
      Exp3ResynthScript ---
      Exp3cResynthScript ---
      Exp1dResynthScript ---
      Exp2dResynthScript ---
      Exp3dResynthScript
  end
  subgraph ResynthFiles[Resynthesized Files]
    direction TB
    Exp1Resynth ---
    Exp2Resynth ---
    Exp3Resynth ---
    Exp3cResynth ---
    Exp1dResynth ---
    Exp2dResynth ---
    Exp3dResynth
  end
  BezierHelpers -- Create curves --> ResynthScripts
  ResynthScripts --> ResynthFiles
end
AvgFiles --> Resynthesis
Resynthesis --> ResynthAnalysis & TcogFigures
ResynthAnalysis -- Cull recordings -->  FixedFiles
```

:::

We can now resynthesize all of the averaged recordings.
The `resynthesis_analysis.qmd` provides visualizations of the resynthesized
files to double check that the resynthesis held. Sometimes rerunning the
resynthesis on the file is good enough, but sometimes there's a voice quality
issue at play.

### Resynthesis Manipulations

Here, "straight" means non-curved onglides and "gradual" means linear trajectories
to the boundary tone target.
The latter contrasts with "bottomout" which has an additional low target to
create an early fall.

 - Exp 1= Monotonal-Gradual. Uses `Scripts/resynthesize_continua.praat`, 
 output saved in `06_ResynthesizedRecordings`
 - Exp 2= LHS-Straight-Gradual. Uses  `Scripts/resynthesize_alignment_continua_exp2.praat`,
 output in `06b_ResynthesizedRecordingsExp2`.
 - Exp 3 = Bitonal-Straight-Gradual. Uses `Scripts/resynthesize_straight_alignment_continua.praat`,
  output in `06c_ResynthesizedRecordingsExp3`.
 - Exp3CURVED = Bitonal-Curved-Gradual. Uses `Scripts/resynthesize_exp3_curved.praat`, 
 output in `06c_ResynthesizedRecordingsExp3CURVED`
 - BottomoutMonotonal = Monotonal-Bottomout. Uses `Scripts/resynthesize_monotonal_bottomout`,
 output in `06d_ResynthesizedRecordingsBottomoutMonotonal`
 - BottomoutLHS = LHS-Curved-Bottomout. Uses `Scripts/resynthesize_LHS_bottomout`,
 output in `06d_ResynthesizedRecordingsBottomoutLHS`
 - BottomoutBitonal = Bitonal-Curved-Bottomout. uses `Scripts/resynthesize_bitonal_bottomout`,
 output in `06d_ResynthesizedR

## Final Normalization Procedures

```{mermaid postprocessing-chart}
flowchart TD
classDef praat fill:#E46E85,stroke:#85144b,color:#ffffff;
classDef python fill:#366994,stroke:#FFC331,color:#FFC331;
classDef rstats fill:#6FA1D0,stroke:#245AAB,color:#ffffff;
classDef shell fill:#111111,stroke:#2ECC40,color:#ffffff;

SilenceNormScript[[norm_silence.praat]]
RMSNormScript[[rms_norm.praat]]
lame[[lame.exe]]

class SilenceNormScript praat
class RMSNormScript praat
class lame shell

FixedFiles["📂07: Reduced Recordings"]
SilNormed["📂08: Silence Duration Normed Recordings"]
RMS["📂09: RMS Normed Recordings"]
FinalFiles["📂10: Final Recordings"]


FixedFiles -- Norm Silence Durations --> SilenceNormScript --> SilNormed
    SilNormed -- RMS Norm  Intensity --> RMSNormScript --> RMS
    RMS -- Convert to MP3 --> lame
    lame --> FinalFiles
```

# Automation

Anything that does not absolutely require human intervention, such as the textgrid annotation at the very beginning, is automated.
This includes running praat scripts, making duration calculations, and running analyses (after they're written of course.)
Automation is done via the `{targets}` R package, which provides makefile-like functionality through R.
You can see the flowchart for the automated steps below, which reruns a target whenever an upstream dependency is modified (eg new file, removed file, edited file).
Through this, we don't have to worry bout rerunning specific scripts in a specific order.

```{r targets-display}
tar_glimpse(targets_only = TRUE)
```

